\section*{Foreword}

Hi reader, the following is a collection of notes I took while taking SFU's MACM
316 (Numerical Analysis) in Spring 2025. Many of the notes are heavily based on
the lecture notes given by Professor Steven Ruuth, which can be found in this
repository at \texttt{./steve's notes}. I have tried to make the notes as
readable as possible, and every file is compiled in a machine-readable format so
you can easily stick them into an AI knowledge base to help you with your
studies.

If you need help getting back up to speed on Linear Algebra, I highly recommend
taking a look at \ulhref{https://venhance.github.io/napkin/Napkin.pdf\#part.4}{An Infinitely Large Napkin} by Venhance. It's a great book that covers all of the
theory of mathematics that you will need to understand the material in this
course.
Overtime, I will be updating these notes to be better structured based on the
chapters of the textbook that the course is structured around.

\subsection*{Current Progress}

Lectures 1 - 4 are cleaned up, corrected and usable to learn off of.
I have separated out the sections
into their appropriate chapters. I am currently working on refactoring the
\enquote{Lecture 00n} files into \enquote{Chapter 00n} files. Next up, I will be
cleaning up Lecture 5.

\subsection*{How to use this book}

Until the notes are perfected, I recommend using them purely as a reference for
studying the material. You can very easily use text search to find any specific
information that you are looking for. When I wrote them, they were not meant to
be used as a textbook, instead, they were meant to be thrown into ChatGPT and
used as a knowledge base to help ChatGPT help you with your studies. Good luck!
